Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2014-01-14 Direction_JINST_v10.tex
the last few versions. --Lindley

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sigma(E)=\sim$5 \% $/\sqrt{E(MeV)}$\cite{kam2013,borexino}. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, this energy resolution is much poorer than other technologies being used for $0\nu\beta\beta$ searches: Ge detectors\cite{gerda2013}, Te bolometers\cite{Alessandria:2011rc}, tracking detectors\cite{SuperNEMO}, liquid Xe time projection chambers (TPCs)\cite{EXO2012} and high pressure gaseous Xe TPCs\cite{NEXTsipm}.

Scintillation light is isotropic. At these low energies, it does not contain sufficient information to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $\sim10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

@@

The Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. The absorption cutoff for the example scintillator above is 370~nm. The index of refraction at  n(370nm)=1.466 and can be translated into an effective Cherenkov threshold for electrons at 0.188~MeV. For a 5~MeV electron, this yields 685 Cherenkov photons per event when integrating from the cutoff wavelength at 360~nm to 550~nm,  the wavelength above which a small number of photons is detected due to a standard bialkali photocathode's quantum efficiency. Lowering the energy to 1~MeV yields 82 Cherenkov photons between 360-550~nm.

@@

\section{Geant4 simulation}
\label{sim_section}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a \GEANT \newline 
\cite{geant4one,geant4two} simulation
has been constructed. The simulation uses \GEANT~version 4.9.6 with the default liquid scintillator
optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene and 1.52~g/l PPO. The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}.  This is a standard scintillator. The attenuation length at 400~nm, the position of the peak standard bialkali photocathod efficiency, is 25~m. The attenuation length drops precipitously at 370~nm from 6.5~m to 0.65~m at the cutoff wavelength at 360~nm.
Variations from the baseline KamLAND case are discussed below.

@@

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. As in the case of optical scattering, reflections at the sphere are a small effect that would create a small tail at longer times. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}. The QE values as a function of wavelength come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

@@

In sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations at 5~MeV. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction. With the successful reconstruction at 5~MeV, we then lower the energy of the simulated electrons and show that it is possible to reconstruct electrons in the range interesting for $0\nu\beta\beta$.

@@

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV; lower energies are studied in the next section.
The electrons are simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0). Figure \ref{fig:reco} (Top)
shows the vertex reconstruction. The vertex is reasonably well-reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in figure \ref{fig:reco} (Top). It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms
used.

@@

We have developed a technique to separate scintillation and Cherenkov light to reconstruct direction of electrons with energies 5~MeV, 2.1~MeV and 1.4~MeV in a liquid scintillator detector. These energies have been chosen to represent typical neutrino-electron elastic scattering energies and
$0\nu\beta\beta$ energies. The Cherenkov threshold for an electron in a typical liquid scintillator is $\sim$0.188~MeV

While scintillation light is isotropic, Cherenkov light with wavelengths above the absorption cutoff $\sim$370~nm carries the information about the direction of the electrons. All light shorter than this gets absorbed and re-emitted isotropically as a part of the scintillation process. On average scintillation light is delayed with respect to the direct Cherenkov light due to chromatic dispersion and and the finite time of the scintillation processes, therefore early light contains directional information.

@@

Reconstruction algorithms developed for water Cherenkov detectors have been applied to this early light. The algorithms are able to converge on reasonable reconstructed vertices and directions for all simulated energies: 5~MeV, 2.1~MeV, 1.4~MeV. As expected, the reduction in photon statistics with lower energies leads to a broadening of the reconstructed vertex by 1.5~cm. Similarly, the mean reconstructed direction drops from XX to XX as the $\cos(\theta)$ distribution broadens.

We have shown that this technique promising, and we plan to continue work on the topic. The ability to reconstruct direction in kiloton-scale scintillation detectors would expand capabilities of neutrino experiments, especially those also searching for neutrino-less double-beta decay. More generally, this technique could be applied wherever scintillation-based detectors are used.

>>>>

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sigma(E)=\sim$5 \% $/\sqrt{E(MeV)}$\cite{kam2013,borexino}. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology for neutrino detectors that can be economically scaled to these large
masses. However, this energy resolution is much poorer than other technologies being used for $0\nu\beta\beta$ searches: Ge detectors\cite{gerda2013}, Te bolometers\cite{Alessandria:2011rc}, tracking detectors\cite{SuperNEMO}, liquid Xe time projection chambers (TPCs)\cite{EXO2012} and high pressure gaseous Xe TPCs\cite{NEXTsipm}.

Scintillation light is isotropic. At these low energies, it does not contain sufficient information to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $\sim10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO,NEXTsipm}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

@@

The Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. The absorption cutoff for the example scintillator above is 370~nm. The index of refraction at  n(370nm)=1.466 and can be translated into an effective Cherenkov threshold for electrons at 0.188~MeV. For a 5~MeV electron, this yields 685 Cherenkov photons per event when integrating from the cutoff wavelength at 370~nm to 550~nm,  the wavelength above which a small number of photons is detected due to a standard bialkali photocathode's quantum efficiency. Lowering the energy to 1~MeV yields 82 Cherenkov photons between 370-550~nm.

@@

\section{Geant4 simulation}
\label{sim_section}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a Geant4\cite{geant4one,geant4two} simulation
has been constructed. The simulation uses Geant4~version 4.9.6 with the default liquid scintillator
optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene and 1.52~g/l PPO. The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}.  This is a standard scintillator. The attenuation length at 400~nm, the position of the peak standard bialkali photocathode efficiency, is 25~m. The attenuation length drops precipitously between 370~nm and 360~nm from 6.5~m to 0.65~m. We use this drop to define the cutoff wavelength at 370~nm. Variations from the baseline KamLAND case are discussed below.

@@

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. As in the case of optical scattering, reflections at the sphere are a small effect that would create a small tail at longer times. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}. The QE values as a function of wavelength come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency. We are neglecting any threshold effects in the photodetector readout electronics.

@@

In sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations at 5~MeV. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the Geant4~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction. With the successful reconstruction at 5~MeV, we then lower the energy of the simulated electrons and show that it is possible to reconstruct electrons in the range interesting for $0\nu\beta\beta$.

@@

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV; lower energies are studied in the next section.
The electrons are simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0). Figure \ref{fig:reco} (Top)
shows the vertex reconstruction. The vertex is reasonably well-reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in figure \ref{fig:reco} (Top). It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms.

@@

We have developed a technique to separate scintillation and Cherenkov light to reconstruct direction of electrons with energies 5~MeV, 2.1~MeV and 1.4~MeV in a liquid scintillator detector. These energies have been chosen to represent typical neutrino-electron elastic scattering energies and
$0\nu\beta\beta$ energies. The Cherenkov threshold for an electron in a typical liquid scintillator is $\sim$0.188~MeV.

While scintillation light is isotropic, Cherenkov light with wavelengths above the absorption cutoff $\sim$370~nm carries the information about the direction of the electrons. All light shorter than this gets absorbed and re-emitted isotropically as a part of the scintillation process. On average scintillation light is delayed with respect to the direct Cherenkov light due to chromatic dispersion and the finite time of the scintillation processes, therefore early light contains directional information.

@@

Reconstruction algorithms developed for water Cherenkov detectors have been applied to this early light. The algorithms are able to converge on reasonable reconstructed vertices and directions for all simulated energies: 5~MeV, 2.1~MeV, 1.4~MeV. As expected, the reduction in photon statistics with lower energies leads to a broadening of the reconstructed vertex by $\sim$1~cm. Similarly, the mean reconstructed direction drops from 0.88 to 0.74 as the $\cos(\theta)$ distribution broadens.

We have shown that this technique is promising, and we plan to continue work on the topic. The ability to reconstruct direction in kiloton-scale scintillation detectors would expand capabilities of neutrino experiments, especially those also searching for neutrino-less double-beta decay. More generally, this technique could be applied wherever scintillation-based detectors are used.
